---
output: md_document
---
# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
```{r, echo=FALSE}
url <- "C:/Users/a/Desktop/DS/5 Reproducible Research/assignment 1/"
setwd(url)
```

File [activity.zip](https://github.com/ua-umka/RepData_PeerAssessment1/blob/master/activity.zip?raw=true) are downloaded and uncompressed in a working directory. To uncompress the file *downloader package* is used.
```{r}
library("downloader")
if (!file.exists("activity.zip")) {
      url <- "https://github.com/ua-umka/RepData_PeerAssessment1/blob/master/activity.zip?raw=true"
      download(url, dest="activity.zip", mode="wb")
      unzip ("activity.zip")
}
data <- read.csv("activity.csv")
```

The data is processed into a format suitable for the analysis by summing up steps for each day. The histogram below shows the result.
```{r, fig.height=4, fig.width=4}
day <- aggregate(steps ~ date, data = data, FUN = sum)
hist(day$steps, main = "Total number of steps taken each day",
     xlab = "Number of steps", col = "blue")
```


## What is mean and median total number of steps taken per day?
```{r}
mean(day$steps)
median(day$steps)
```


## What is the average daily activity pattern?
To address this question, a time series plot is drawn. X-axis represents the 5-minute interval, while y-axis - the average number of steps taken during those intervals.
```{r}
interv <- aggregate(steps ~ interval, data = data, mean)
plot(interv$interval, interv$steps, type = "l", xlab = "Interval", ylab = "Steps", main = "Average number of steps by intervals")
```

**Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**


## Imputing missing values
First, we'll calculate the total number of rows with at least one missing value in it.
```{r}
table(complete.cases(data[[1]], data[[2]], data[[3]]))[[1]]
```
Let's check how many NAs are in the steps column.
```{r}
sum(is.na(data$steps))
```
The same number, so all NAs are in the step column.

#### Filling in the NAs
The strategy for filling in the missing values in the dataset will be to fill them whith the mean for an appropriate 5-minute interval calculated before.

Let's create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
dataWhNa <- data
row <- which(is.na(data))
for (i in row) {
      dataWhNa[i,1] <- interv[interv$interval == dataWhNa[i,3], 2]
}
```
Let's check the difference between the old one and the new one without NAs by drawing them side by side.
```{r}
par(mfrow = c(1,2))
hist(day$steps, main = "With NAs",
     xlab = "Number of steps", col = "blue")
dayWhNas <- aggregate(steps ~ date, data = dataWhNa, FUN = sum)
hist(day$steps, main = "Without NAs",
     xlab = "Number of steps", col = "blue")
```

Doesn't see any difference. How about mean and median of the total number of steps taken per day?

```{r}
mean(dayWhNas$steps); median(dayWhNas$steps)
# The difference:
mean(dayWhNas$steps) - mean(day$steps); median(day$steps) - median(dayWhNas$steps)
```

So there is a slight change in median. The median of the data without NAs is higher by 1.19 step per day.

## Are there differences in activity patterns between weekdays and weekends?
Let's create a new factor variable in the dataset with two levels Weekend and Weekday.
```{r}
weekday <- character()
for (i in 1:(dim(dataWhNa)[1])) {
      weekday[i] <-  weekdays(as.Date(dataWhNa$date[i]))
}
weekday[weekday %in% c("Saturday", "Sunday")] <- "Weekend"
weekday[weekday != "Weekend"] <- "Weekday"
dataWhNa <- cbind(dataWhNa, weekday)
weekDays <- dataWhNa[dataWhNa$weekday == "Weekday",]
weekEnds <- dataWhNa[dataWhNa$weekday == "Weekend",]
```

Now when we have the factor in the data, let's look at the plots of the average number of steps separately for weekdays and weekends. 
```{r}
weekDays <- aggregate(steps ~ interval, data = weekDays, mean)
weekEnds <- aggregate(steps ~ interval, data = weekEnds, mean)
par(mfrow = c(1,2))
plot(weekDays, type = "l", main = "Weekdays", ylim = c(0,250), col = "blue")
plot(weekEnds, type = "l", main = "Weekends", ylim = c(0,250), col = "red")

```

They definitely differs. We don't see such a big step spikes during weekends as we see for Weekdays, but overall the are more steps during weekends than weekdays.